All the concepts behind TrueSkill Through Time: the \proglang{Julia}, \proglang{Python} and \proglang{R} packages (Draft)

### 2020-12-11 article.tex
Paper TTT: revisi'on de intro con Gus

\en{Humans develop complex skills because of an special integration of biological, cognitive and social processes~\citep{Koster2020}.}
\es{Los humanos desarrollan habilidades complejas gracias a una integraci\'on especial de los procesos biol\'ogicos, cognitivos y sociales~\citep{Koster2020}.}
\en{Our exceptional cognitive ability to imitate, combined with human long periods of juvenile dependency and postreproductive life span, allows us to learn things from others and transmit innovations through generations~\citep{Richerson2020}.}
\es{Nuestra extraordinaria capacidad para imitar, combinada con los largos períodos de aprendizaje juvenil y vida posreproductiva, permite a los humanos aprender de los dem\'as y transmitir las innovaci\'on a trav\'es de la generaciones~\citep{citar}.}
%
\en{As a population-based process, human adaptation is also affected by demographic characteristics, such as the size and structure of populations~\citep{Derex2020}.}
\es{Al ser un proceso poblacional, la adaptaci\'on humana tambi\'en se ve afectada por caracter\'isticas demogr\'aficas, como el tama\~no y estructura de las poblaciones~\citep{Derex2020}.} 
% \en{Understanding the factors impacting in the positive or negative changes of skill could be used to develop new training strategies or adapting the environment to avoid unnecessary tasks.}
% \es{Hacer Conocer c\'omo cambian la habilidades de la personas en el tiempo es crucial en muchas \'areas.}

\en{Knowing how individual skills change over time is essential in sports and educational contexts.}
\es{Conocer c\'omo cambian las habilidades individuales a lo largo del tiempo es esencial para en contextos deportivos y educativos.}
%
\en{Since they are hidden variables, the best we can do is estimate them based on their direct observable consequences: the outcome of problem solving and competitions.}
\es{Dado que son variables ocultas, lo mejor que podemos hacer es estimarlas a partir de sus consecuencias observables directas: el producto de resoluci\'on de problemas y competencias.}
%
\en{Considering the frequency of positive results as an indicator of the ability of individuals is a misconception, mainly because its value also depends on the difficulty of the challenges.}
\es{Considerar la frecuencia de resultados positivos como indicador de la habilidad de los individuos es un error de concepto, fundamentalmente porque su valor depende tambi\'en de la dificultad de los desaf\'ios.}
%
\en{For this reason, all widely used skill estimators are based on pairwise comparisons.}
\es{Por esta raz\'on, todos los estimadores de habilidad ampliamente usados se basa en comparaciones por pares.}
%
\en{Since the first generative models, proposed almost a century ago by~\cite{Thurstone1927} and~\cite{zermelo1929-translation}, it is assumed that the probability of an observed result, $r$, depends on the performance, $p$, of the agent $i$ and their opponent $j$, expressed as $P(\, r \,|\, p_i, \, p_j \,)$.}
\es{Desde los primeros modelos generativos, propuestos hace casi un siglo por~\cite{Thurstone1927} y~\cite{zermelo1929-translation}, se supone que la probabilidad de un resultado observado, $r$, depende del rendimiento, $p$, del agente $i$ y de su oponente $j$, expresada como $P(\, r \,|\, p_i, \, p_j \,)$.}
%
\en{The field continued to progress with the work of~\cite{Bradley1952} and~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}.}
\es{El campo sigui\'o progresando con los trabajos de \cite{Bradley1952} y~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}.}
% 
\en{A breakthrough took place with the methodology developed by~\cite{Elo2008} for the US Chess Federation (USCF), which is still used by the International Chess Federation (FIDE).}
\es{Un gran avance produjo la metodolog\'ia desarrollada por~\cite{Elo2008} para la Federaci\'on de Ajedrez de los Estados Unidos (USCF), adoptada hasta el d\'ia de hoy por la Federaci\'on Internacional de Ajedrez (FIDE).}

@@

\en{All currently widely used skill estimators share some variant of the probabilistic model proposed by the Elo~\citep{Glickman1999,Herbrich2007,VanDerLinden2016,Fox2010}.}
\es{Todos los estimadores de habilidad ampliamente utilizados actualmente comparten alguna variante del modelo probabilístico propuesto por el Elo~\cite{Glickman1999, Herbrich2007, VanDerLinden2016, Fox2010}.}
%
\begin{figure}[h!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r_{ij}$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r_{ij} = (d_{ij}>0)$};

    \node[latent, above=of r, yshift=-0.45cm] (d) {$d_{ij}$} ; %
    \node[const, right=of d] (dd) {\normalsize $ d_{ij}=p_i-p_j$}; 
    \node[const, left=of d, xshift=-1.35cm] (d_name) {\small \en{Difference}\es{Diferencia}:};

@@

    \edge {d} {r};
    \edge {p1,p2} {d};
    \edge {s1} {p1};
    \edge {s2} {p2};
    %\node[invisible, right=of p2, xshift=4.35cm] (s-dist) {};
}
     \caption{
     \en{Generative model in which skills cause the observable results mediated by the difference of hidden performances.}
    \es{Modelo generativo en el que las habilidades causan los resultados observables a trav\'es de la diferencia de rendimientos ocultos.}
    %
    \en{At each event, agents exhibit a random performance around their unknown true skill, $p \sim \N(s,\beta^2)$.}
    \es{En cada evento, los agentes exhiben una rendimiento aleatorio centrado en la verdadera y desconocida habilidad, $p \sim \N(s,\beta^2)$.}
    %
    \en{The one with the highest performance wins, $r_{ij} = p_i > p_j$.}
    \es{Gana quien haya obtenido mayor rendimiento, $r_{ij} = p_i > p_j$.}
    %
    \en{This graphical representation defines a joint probability distribution.}
    \es{Esta representaci\'on gr\'afica define una distribuci\'on de probabilidad conjunta.}
    %
    \en{Observable variables are painted gray, hidden in white, and constants are shown as black dots.}
    \es{Las variables observables se pintan de gris, la ocultas en blanco, y las constantes se muestran como puntos negros.}
    }
    \label{fig:generative_model}
\label{modelo}
\end{figure}
%
%
\en{Figure~\ref{fig:generative_model} provides a graphical representation in which skills generates the observable result.}
\es{La figura~\ref{fig:generative_model} ofrece una representaci\'on gr\'afica en la que las habilidades generan el resultado observable.}
%
\en{The model assumes that the result is a direct consequence of the performances, which vary between events around the true skill with some constant noise,  $\N(p\,|\,s,\beta^2)$.}
\es{El modelo supone que el resultado es consecuencia directa de los rendimientos, que varían entre eventos alrededor de la verdadera habilidad con cierto ruido constante, $\N(p\,|\,s,\beta^2)$.}
%
% \en{Then, the observed result is a direct consequence of the difference of the performance of each agent in a particular event (i.e. game), $d_{ij}=p_i - p_j$.}
%
\en{The one with the highest performance wins, $r = (p_i > p_j)$.}
\es{Gana quien haya obtenido mayor rendimiento, $r = (p_i > p_j)$.}

@@

\en{Elo's methodological solution is simple and astute: update the previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, based on the surprise, i.e. the complement of the prediction of the observed result.}
\es{La soluci\'on metodol\'ogica de Elo es simple y astuta: actualizar las estimaciones previas, $s_{i_\text{old}}$ y $s_{j_\text{old}}$, en base a la sorpresa, i.e. el complemento de la predicci\'on del resultado observado.}
%
%empezar con estimaciones arbitrarias
%
\begin{equation} \label{eq:elo_delta}
 \Delta = \underbrace{\left(1-P(\,r\,|\,s_{i_\text{old}},s_{j_\text{old}}\,)\right)}_{\text{\en{Surprise}\es{Sorpresa}}}
\end{equation}
%
\en{Where the probability arises from instantiating the previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, in the generative model (details in section~\ref{sec:exactSolution}).}
\es{Donde la probabilidad surge de instanciar las estimaciones previas en el modelo generativo (detalles en la sección~\ref{sec:exactSolution}).}
%
\en{The idea is that the surprise, $\Delta$, is related to the quality of the previous estimates and therefore serves to update them.}
\es{La idea es que la sorpresa, $\Delta$, está relacionada con la calidad de las estimaciones previas y por lo tanto sirvan para actualizarlas. }
%
\en{Unexpected results would indicate that current estimates are not entirely correct and should be updated to a greater extent than if they had occurred as expected.}
\es{Los resultados inesperados indicar\'ian que las estimaciones actuales no son del todo correctas y deber\'ian actualizarse en mayor medida que si hubieran ocurrido como se esperaba.}
%
\begin{equation}\label{eq:elo_update}
 s_{\text{winner}_\text{new}} = s_{\text{winner}_\text{old}} + \Delta \ \ \ \ \ s_{\text{loser}_\text{new}} = s_{\text{loser}_\text{old}} - \Delta 
\end{equation}
%
\en{Where the surprise, $\Delta$, acts as a correction factor for both previous estimates.}
\es{Donde la sopresa, $\Delta$, act\'ua como factor de correcci\'on para ambas estimaciones previas.}
%
\en{This procedure can recover the relative scale of the agents, starting from arbitrary initial values.}
\es{Esta soluci\'on puede recuperar la escala relativa de los agentes, partiendo de de valores iniciales arbitrarios.}
%
\en{However, it does have a major weaknesses.}
\es{Sin embargo, tiene algunas debilidades importantes.}
%
\en{The update rule (Eq.~\ref{eq:elo_update}) is symmetric, so what one agent wins is lost by the other.}
\es{La regla de actualizaci\'on (Eq.~\ref{eq:elo_update}) es sim\'etrica, as\'i que lo que un agente gana lo pierde el otro.}
%
\en{Because new agents start with arbitrary estimates (the same initial value for any individual), they tend to generate high surprise and abruptly modify estimates that had already converged correctly.}
\es{Debido a que a los agentes nuevos comienzan con estimaciones arbitrarias (el mismo valor inicial para cualquier individuo), ellas tienden a generan alta sorpresa y a modificar bruscamente estimaciones que ya habían convergido correctamente.}
%
\en{The weak point of the methodological solution proposed by Elo is the lack of a notion of uncertainty.}
\es{El punto débil de la solución metodológica propuesta por Elo es la falta de una noción de incertidumbre.}
%
\en{To solve this problem, an ad-hoc solution was proposed: reducing the impact of the surprise, $\Delta$, based on the number of times the agent has participated.}
\es{Para resolver este problema, una soluci\'on ad-hoc fue propuesta: reducir el impacto de la sorpresa, $\Delta$, en funci\'on de la cantidad de veces que el agente ha participado.}
%
\en{This is the role played by the K-factor used by FIDE,  $\Delta_i = \Delta \cdot K_i$.}
\es{Ese es rol que desempe\~na el K-factor usado por la FIDE, $\Delta_i = \Delta \cdot K_i$.}

@@

\en{Instead of selecting a single value as an estimate, we would like to quantify the ``relative certainty'' associated with each of the skill hypotheses.}
\es{En vez seleccionar un único valor como estimación, quisiéramos cuantificar la ``certidumbre relativa'' asociada a cada una de las hipótesis de habilidad.}
\en{The use of probability theory to handle this type of uncertainty (or beliefs), known as Bayesian inference, has been widely adopted not only because it has been shown to be a successful tool in practice~\citep{bishop2006-PRML}, but also because its rules have been proven to be the only ones that guarantee consistent reasoning~\citep{jaynes2003-bookProbabilityTheory,VanHorn2003}.}
\es{El uso de la teor\'ia de la probabilidad para tratar este tipo de incertidumbre (o creencias), conocida como inferencia Bayesiana, se ha adoptado ampliamente no sólo porque ha mostrado ser una herramienta exitosa en la práctica \citep{bishop2006-PRML}, sino también porque se ha demostrado que sus reglas son las únicas que garantizan un razonamiento consistente~\citep{jaynes2003-bookProbabilityTheory,VanHorn2003}.}

@@

@@

\en{Where $ P (\cdot) $ and $ p (\cdot) $ represent discrete and continuous probabilities distributions respectively.}

@@

\en{The \ref{eq:sum_rule} state that any marginal distribution can be obtained by integrating, or summing up, the joint distribution.}

@@

\en{Additionally, the \ref{eq:product_rule} state that any joint distribution can be expressed as the product of one-dimensional conditional distributions.}
\en{From these rules, jointly with the symmetry property $p(x,y) = p(y,x)$, we can immediately obtain the~\ref{eq:bayes_theorem}:}

@@

\en{In our case, to reduce the uncertainty about the hidden skill using the information provided by the observed result and the described causal model, we need to solve the following equation.}
\es{En nuestro caso, para reducir la incertidumbre sobre la habilidad oculta usando la informaci\'on que nos ofrece el resultado observado y el modelo causal descrito, necesitamos resolver la siguiente ecuaci\'on.}

@@

\en{Where the hidden skills are the hypotheses, the observed results are the data, and the model is defined by the figure~\ref{fig:generative_model}.}

@@

\es{Por la forma del posterior exacto, una Gaussiana parece una aproximaci\'on suficientemente buena, que nos permitiría evitar el costo computacional de las metodolog\'ias de sampleo.}

@@

\es{Esta clase de m\'etodos, que minizan la divergencia Kullback-Leibler entre la distribuci\'on verdadera y aproximada, permiten realizar Inferencia Bayesiana en situaciones que de otra forma no sería posibles.}

@@

\es{Debido a que el enfoque de filtrado no se surge de ningún modelo probabilístico, sufre de un serie de problemas.}

@@

% impidiendo así la comparabilidad temporal y entre comunidades.

@@

      }  
  }
  \caption{
  \en{Schematic representation of a single causal model for the entire history of events.}
  \es{Representaci\'on esquemática de un único modelo causal para toda la historia de eventos.}
  %
  \en{Only the neighbors of a sequence of estimates are displayed.}
  \es{S\'olo se muestra los vecinos de una secuencia de estimaciones.}
  %
%   \en{Its probabilistic solution generates the approach known as \emph{smoothing}, in which historical information flows through the whole system.}
%   \es{La soluci\'on probabilística da como resultado el abordaje conocido como \emph{smoothing}, en el que la informaci\'on hist\'orica fluye por todo el sistema.}
%   %
%   %\en{These types of models are known as \emph{state-space models}.}
%   %\es{Este tipo de modelos se conocen como \emph{state space models}.}
  }
  \label{fig:smoothing}
\end{figure}
%
\en{By applying the rules of probability to this temporal model, historical information naturally propagate throughout the system, solving the problems of the filtering approach.}
\es{Al aplicar las reglas de la probabilidad sobre este modelo temporal, la informaci\'on hist\'orica propaga naturalmente hacia todo el sistema, resolviendo los problemas del enfoque de filtro.}
%
\en{A prior distribution at an event $t$, with $\gamma = 0$, is now the product of all past and future information.}
\es{La distribuci\'on a prior en un evento $t$, con $\gamma = 0$, es ahora la multiplicaci\'on de toda la informaci\'on pasada y futura.}
%
\begin{equation}\label{eq:smooth_prior}
 \text{Prior}_{i_t} = \text{Prior}_{i_0} \underbrace{\prod_{k = 1}^{t-1} \text{\en{Likelihood}\es{Verosimilitud}}_{i_k}}_{\text{\en{Past information}\es{Informaci\'on pasada}}} \underbrace{\prod_{k = t + 1}^{T} \text{\en{Likelihood}\es{Verosimilitud}}_{i_k}}_{\text{\en{Future information}\es{Informaci\'on futura}}}
\end{equation}
%
\en{This generates a mutual dependency between likelihoods, forcing us to implement an eficiente iterative algorithm to solve inference (details at secci\'on~\ref{sec:throguthTime})}
%
\es{Esto genera una mutua dependencia entre verosimilitudes, obligándonos a implementar un algoritmo para resolver la inferencia (detalles en la secci\'on~\ref{sec:throguthTime}).}
%
\en{The loopy belief propagation is procedure that converge with a few linear iterations over the data, scaling to millions of observations in a few minutes.}
\es{Este procedimiento converge con unas pocas iteraciones lineales sobre los datos, escalando a millones de observaciones en pocos minutos.}
%
\en{At the beginning, the likelihoods that are not yet defined are initialized with a neutral value or simply excluded.}
\es{Al empezar, las verosimilitudes que aun no están definidas se inicializan con un valor neutro o simplemente se las excluy\'e.}
%
\en{In Figure~\ref{fig:smooth_example} we show the behavior of the estimates in a data set with two agents and two events.}
\es{En la figura~\ref{fig:smooth_example} se muestra el comportamiento de las estimaciones en conjunto de datos de dos agentes y dos eventos.}
%
\begin{figure}[h!]
  \centering
  \scalebox{.9}{
    \tikz{ %
      \node[latent] (s10) {$s_{a_0}$} ;
      %
      \node[latent,  below=of s10,yshift=-1cm] (s11) {$s_{a_1}$} ;

@@

\en{The efficiency of TrueSkill Throguth Time is due to the fact that margins, whether exact or approximate, are solved analytically.}
\es{La eficiencia de TrueSkill Throguth Time se debe a que las marginales, sean exactas o aproximadas, se resuelven de forma analítica.}
%
\en{In this section we list the properties that we will use to derive the exact messages that arise from the sum-product algorithm.}
\es{En esta secci\'on enumeramos las propiedades que usaremos para derivar los mensajes exactos que surgen del \emph{sum-product algorithm}.}
%
\en{Although these properties are widely known, we attach their full demonstrations in the supplemental material.}
\es{Aunque estas propiedades son ampliamente conocidas, adjuntamos sus demostraciones completas en material suplementario.}

@@

\es{El siguiente mensaje, que la variable $s_i$ le envía al factor rendimiento $f_{p_i}$ tambi\'en será el prior.}

@@

\es{Recordar que los mensajes que envían las variables es tan solo la multiplicaci\'on de los mensajes que esa variable recibe del resto de los factores.}

@@

\es{Por ser de trivial soluci\'on, no haremos explícitos este tipo de mensajes.}

>>>>

\en{Humans develop complex skills because of an special integration of biological, cognitive and social processes~\citep{Koster2020}.}
\es{Los humanos desarrollan habilidades complejas gracias a una integraci\'on especial de los procesos biol\'ogicos, cognitivos y sociales~\citep{Koster2020}.}
\en{Our exceptional cognitive ability to imitate, combined with human long periods of juvenile dependency and postreproductive life span, allows us to learn things from others and transmit innovations through generations~\citep{Richerson2020}.}
\es{Nuestra extraordinaria capacidad para imitar, combinada con los largos per\'iodos de aprendizaje juvenil y vida posreproductiva, permite a los humanos aprender de los dem\'as y transmitir las innovaci\'on a trav\'es de la generaciones~\citep{citar}.}
%
\en{As a population-based process, human adaptation is also affected by demographic characteristics, such as the size and structure of populations~\citep{Derex2020}.}
\es{Al ser un proceso poblacional, la adaptaci\'on humana tambi\'en se ve afectada por caracter\'isticas demogr\'aficas, como el tama\~no y estructura de las poblaciones~\citep{Derex2020}.} 
% \en{Understanding the factors impacting in the positive or negative changes of skill could be used to develop new training strategies or adapting the environment to avoid unnecessary tasks.}
% \es{Hacer Conocer c\'omo cambian la habilidades de la personas en el tiempo es crucial en muchas \'areas.}

\en{Knowing how individual skills change over time is essential in sports and educational contexts.}
\es{Conocer c\'omo cambian las habilidades individuales a lo largo del tiempo es esencial para en contextos deportivos y educativos.}
%
\en{Since they are hidden variables, the best we can do is to estimate them based on their direct observable consequences: the outcome of problem solving and competitions.}
\es{Dado que son variables ocultas, lo mejor que podemos hacer es estimarlas a partir de sus consecuencias observables directas: el producto de resoluci\'on de problemas y competencias.}
%
\en{Only considering the frequency of positive results as an indicator of the individuals' ability could lead to wrong approximations, mainly because the outcome also depends on the difficulty of the challenge.}
\es{Considerar la frecuencia de resultados positivos como indicador de la habilidad de los individuos es un error de concepto, fundamentalmente porque su valor depende tambi\'en de la dificultad de los desaf\'ios.}
%
\en{For this reason, all widely used skill estimators are based on pairwise comparisons.}
\es{Por esta raz\'on, todos los estimadores de habilidad ampliamente usados se basa en comparaciones por pares.}
%
\en{Since the first generative models, proposed almost a century ago by~\cite{Thurstone1927} and~\cite{zermelo1929-translation}, it is assumed that the probability of an observed result, $r$, depends on the performance, $p$, of the agent $i$ and their opponent $j$, expressed as $P(\, r \,|\, p_i, \, p_j \,)$.}
\es{Desde los primeros modelos generativos, propuestos hace casi un siglo por~\cite{Thurstone1927} y~\cite{zermelo1929-translation}, se supone que la probabilidad de un resultado observado, $r$, depende del rendimiento, $p$, del agente $i$ y de su oponente $j$, expresada como $P(\, r \,|\, p_i, \, p_j \,)$.}
%
\en{The field continued to progress with the work of~\cite{Bradley1952} and~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c} leading to the breakthrough that took place with the methodology developed by~\cite{Elo2008} for the US Chess Federation (USCF), which is still used by the International Chess Federation (FIDE).}

\es{El campo sigui\'o progresando con los trabajos de \cite{Bradley1952} y~\cite{Mosteller1951a,Mosteller1951b,Mosteller1951c}.}
% 
\es{Un gran avance produjo la metodolog\'ia desarrollada por~\cite{Elo2008} para la Federaci\'on de Ajedrez de los Estados Unidos (USCF), adoptada hasta el d\'ia de hoy por la Federaci\'on Internacional de Ajedrez (FIDE).}

@@

\en{All currently widely used skill estimators share some variant of the probabilistic model proposed by the Elo~\citep{Glickman1999,Herbrich2007,VanDerLinden2016,Fox2010}.}
\es{Todos los estimadores de habilidad ampliamente utilizados actualmente comparten alguna variante del modelo probabil\'istico propuesto por el Elo~\cite{Glickman1999, Herbrich2007, VanDerLinden2016, Fox2010}.}
%
\begin{figure}[h!]
\centering \small
    \tikz{         
    \node[det, fill=black!10] (r) {$r$} ; 
    \node[const, left=of r, xshift=-1.35cm] (r_name) {\small \en{Result}\es{Resultado}:}; 
    \node[const, right=of r] (dr) {\normalsize $ r = (d > 0)$};

    \node[latent, above=of r, yshift=-0.45cm] (d) {$d$} ; %
    \node[const, right=of d] (dd) {\normalsize $ d = p_i-p_j$}; 
    \node[const, left=of d, xshift=-1.35cm] (d_name) {\small \en{Difference}\es{Diferencia}:};

@@

    \edge {d} {r};
    \edge {p1,p2} {d};
    \edge {s1} {p1};
    \edge {s2} {p2};
    %\node[invisible, right=of p2, xshift=4.35cm] (s-dist) {};
}
     \caption{
     \en{Generative model in which skills cause the observable results mediated by the difference of hidden performances.}
    \es{Modelo generativo en el que las habilidades causan los resultados observables a trav\'es de la diferencia de rendimientos ocultos.}
    %
    \en{At each event, agents exhibit a random performance around their unknown true skill, $p \sim \N(s,\beta^2)$.}
    \es{En cada evento, los agentes exhiben una rendimiento aleatorio centrado en la verdadera y desconocida habilidad, $p \sim \N(s,\beta^2)$.}
    %
    \en{The one with the highest performance wins, $r = p_i > p_j$.}
    \es{Gana quien haya obtenido mayor rendimiento, $r = p_i > p_j$.}
    %
    \en{This graphical representation defines a joint probability distribution.}
    \es{Esta representaci\'on gr\'afica define una distribuci\'on de probabilidad conjunta.}
    %
    \en{Observable variables are painted gray, hidden in white, and constants are shown as black dots.}
    \es{Las variables observables se pintan de gris, la ocultas en blanco, y las constantes se muestran como puntos negros.}
    }
    \label{fig:generative_model}
\label{modelo}
\end{figure}
%
%
\en{Figure~\ref{fig:generative_model} provides a graphical representation in which skills generates the observable result.}
\es{La figura~\ref{fig:generative_model} ofrece una representaci\'on gr\'afica en la que las habilidades generan el resultado observable.}
%
\en{The model assumes that the result is a direct consequence of the performances, which vary between events around the true skill with some constant noise,  $\N(p\,|\,s,\beta^2)$.}
\es{El modelo supone que el resultado es consecuencia directa de los rendimientos, que var\'ian entre eventos alrededor de la verdadera habilidad con cierto ruido constante, $\N(p\,|\,s,\beta^2)$.}
%
% \en{Then, the observed result is a direct consequence of the difference of the performance of each agent in a particular event (i.e. game), $d_{ij}=p_i - p_j$.}
%
\en{The one with the highest performance wins, $r = (p_i > p_j)$.}
\es{Gana quien haya obtenido mayor rendimiento, $r = (p_i > p_j)$.}

@@

\en{Elo's methodological solution is simple and smart: update both previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, based on the surprise (i.e. the complement of the prediction of the observed result).}
\es{La soluci\'on metodol\'ogica de Elo es simple y astuta: actualizar las estimaciones previas, $s_{i_\text{old}}$ y $s_{j_\text{old}}$, en base a la sorpresa, i.e. el complemento de la predicci\'on del resultado observado.}
%
%empezar con estimaciones arbitrarias
%
\begin{equation} \label{eq:elo_delta}
 \Delta = \underbrace{\left(1-P(\,r\,|\,s_{i_\text{old}},s_{j_\text{old}}\,)\right)}_{\text{\en{Surprise}\es{Sorpresa}}}
\end{equation}
%
\en{where the probability arises from instantiating the previous estimates, $s_{i_\text{old}}$ and $s_{j_\text{old}}$, in the generative model (details in section~\ref{sec:exactSolution}).}
\es{Donde la probabilidad surge de instanciar las estimaciones previas en el modelo generativo (detalles en la sección~\ref{sec:exactSolution}).}
%
\en{The idea is that the surprise, $\Delta$, is related to how good are the previous estimates and therefore serves to update them.}
\es{La idea es que la sorpresa, $\Delta$, está relacionada con la calidad de las estimaciones previas y por lo tanto sirvan para actualizarlas. }
%
\en{Unexpected results would indicate that current estimates should be updated to a greater extent than if they had occurred as expected.}
\es{Los resultados inesperados indicar\'ian que las estimaciones actuales no son del todo correctas y deber\'ian actualizarse en mayor medida que si hubieran ocurrido como se esperaba.}
%
\begin{equation}\label{eq:elo_update}
 s_{\text{winner}_\text{new}} = s_{\text{winner}_\text{old}} + \Delta \ \ \ \ \ s_{\text{loser}_\text{new}} = s_{\text{loser}_\text{old}} - \Delta 
\end{equation}
%
\en{where the surprise $\Delta$ acts as a correction factor for both previous estimates.}
\es{Donde la sopresa, $\Delta$, act\'ua como factor de correcci\'on para ambas estimaciones previas.}
%
\en{This procedure can recover the relative scale of the agents, starting from arbitrary initial values.}
\es{Esta soluci\'on puede recuperar la escala relativa de los agentes, partiendo de de valores iniciales arbitrarios.}
%
\en{However, it does have a major weaknesses.}
\es{Sin embargo, tiene algunas debilidades importantes.}
%
\en{The update rule (Eq.~\ref{eq:elo_update}) is symmetric, so what one agent wins is lost by the other.}
\es{La regla de actualizaci\'on (Eq.~\ref{eq:elo_update}) es sim\'etrica, as\'i que lo que un agente gana lo pierde el otro.}
%
\en{Because new agents start with arbitrary estimates (the same initial value is used for all individuals), they tend to generate high surprise and abruptly modify estimates that had already converged.}
\es{Debido a que a los agentes nuevos comienzan con estimaciones arbitrarias (el mismo valor inicial para cualquier individuo), ellas tienden a generan alta sorpresa y a modificar bruscamente estimaciones que ya hab\'ian convergido.}
%
\en{The weak point of the methodological solution proposed by Elo is the lack of the notion of uncertainty.}
\es{El punto débil de la solución metodológica propuesta por Elo es la falta de una noción de incertidumbre.}
%
\en{To solve this problem, an ad-hoc solution was proposed: reducing the impact of the surprise based on the number of times the agent has participated previously.}
\es{Para resolver este problema, una soluci\'on ad-hoc fue propuesta: reducir el impacto de la sorpresa, $\Delta$, en funci\'on de la cantidad de veces que el agente ha participado.}
%
% \en{This is the role played by the K-factor used by FIDE,  $\Delta_i = \Delta \cdot K_i$.}
% \es{Ese es rol que desempe\~na el K-factor usado por la FIDE, $\Delta_i = \Delta \cdot K_i$.}

@@

\en{Instead of selecting a single value as an estimate, we would like to quantify the ``relative certainty'' associated with each of the skill hypotheses.}
\es{En vez seleccionar un único valor como estimación, quisiéramos cuantificar la ``certidumbre relativa'' asociada a cada una de las hipótesis de habilidad.}
%
\en{The use of probability theory to handle this type of uncertainty (or beliefs), known as Bayesian inference, has been widely adopted not only because it has been shown to be a successful tool in practice~\citep{bishop2006-PRML}, but also because its rules have been proven to be the only ones that guarantee consistent reasoning~\citep{jaynes2003-bookProbabilityTheory,VanHorn2003}.}
\es{El uso de la teor\'ia de la probabilidad para tratar este tipo de incertidumbre (o creencias), conocida como inferencia Bayesiana, se ha adoptado ampliamente no sólo porque ha mostrado ser una herramienta exitosa en la práctica \citep{bishop2006-PRML}, sino también porque se ha demostrado que sus reglas son las únicas que garantizan un razonamiento consistente~\citep{jaynes2003-bookProbabilityTheory,VanHorn2003}.}

@@

\en{The \ref{eq:sum_rule} states that any marginal distribution can be obtained by integrating, or summing up, the joint distribution:}

@@

\en{where $ P (\cdot) $ and $ p (\cdot) $ represent discrete and continuous probabilities distributions respectively.}

@@

\en{Additionally, the \ref{eq:product_rule} states that any joint distribution can be expressed as the product of one-dimensional conditional distributions.}

@@

\en{From these rules and jointly with the symmetry property $p(x,y) = p(y,x)$, we can immediately obtain the~\ref{eq:bayes_theorem}:}

@@

\en{To reduce the uncertainty about the hidden skill using the information provided by the observed result and the described causal model, we need to solve:}
\es{En nuestro caso, para reducir la incertidumbre sobre la habilidad oculta usando la informaci\'on que nos ofrece el resultado observado y el modelo causal descrito:}

@@

\en{where the hidden skills are the hypotheses, the observed results are the data, and the model is defined by the figure~\ref{fig:generative_model}.}

@@

\es{Por la forma del posterior exacto, una Gaussiana parece una aproximaci\'on suficientemente buena, que nos permitir\'ia evitar el costo computacional de las metodolog\'ias de sampleo.}

@@

\es{Esta clase de m\'etodos, que minizan la divergencia Kullback-Leibler entre la distribuci\'on verdadera y aproximada, permiten realizar Inferencia Bayesiana en situaciones que de otra forma no ser\'ia posibles.}

@@

\es{Debido a que el enfoque de filtrado no se surge de ningún modelo probabil\'istico, sufre de un serie de problemas.}

@@

% impidiendo as\'i la comparabilidad temporal y entre comunidades.

@@

      }  
  }
  \caption{
  \en{Schematic representation of a single causal model for the entire history of events.}
  \es{Representaci\'on esquemática de un único modelo causal para toda la historia de eventos.}
  %
  \en{Only the neighbors of a sequence of estimates are displayed.}
  \es{S\'olo se muestra los vecinos de una secuencia de estimaciones.}
  %
%   \en{Its probabilistic solution generates the approach known as \emph{smoothing}, in which historical information flows through the whole system.}
%   \es{La soluci\'on probabil\'istica da como resultado el abordaje conocido como \emph{smoothing}, en el que la informaci\'on hist\'orica fluye por todo el sistema.}
%   %
%   %\en{These types of models are known as \emph{state-space models}.}
%   %\es{Este tipo de modelos se conocen como \emph{state space models}.}
  }
  \label{fig:smoothing}
\end{figure}
%
\en{By applying the rules of probability to this temporal model, historical information naturally propagate throughout the system, solving the problems of the filtering approach.}
\es{Al aplicar las reglas de la probabilidad sobre este modelo temporal, la informaci\'on hist\'orica propaga naturalmente hacia todo el sistema, resolviendo los problemas del enfoque de filtro.}
%
\en{A prior distribution at an event $t$, with $\gamma = 0$, is now the product of all past and future information.}
\es{La distribuci\'on a prior en un evento $t$, con $\gamma = 0$, es ahora la multiplicaci\'on de toda la informaci\'on pasada y futura.}
%
\begin{equation}\label{eq:smooth_prior}
 \text{Prior}_{i_t} = \text{Prior}_{i_0} \underbrace{\prod_{k = 1}^{t-1} \text{\en{Likelihood}\es{Verosimilitud}}_{i_k}}_{\text{\en{Past information}\es{Informaci\'on pasada}}} \underbrace{\prod_{k = t + 1}^{T} \text{\en{Likelihood}\es{Verosimilitud}}_{i_k}}_{\text{\en{Future information}\es{Informaci\'on futura}}}
\end{equation}
%
\en{This generates a mutual dependency between likelihoods, forcing us to implement an eficiente iterative algorithm to solve inference (details at secci\'on~\ref{sec:throguthTime})}
%
\es{Esto genera una mutua dependencia entre verosimilitudes, obligándonos a implementar un algoritmo para resolver la inferencia (detalles en la secci\'on~\ref{sec:throguthTime}).}
%
\en{The loopy belief propagation is procedure that converge with a few linear iterations over the data, scaling to millions of observations in a few minutes.}
\es{Este procedimiento converge con unas pocas iteraciones lineales sobre los datos, escalando a millones de observaciones en pocos minutos.}
%
\en{At the beginning, the likelihoods that are not yet defined are initialized with a neutral value or simply excluded.}
\es{Al empezar, las verosimilitudes que aun no están definidas se inicializan con un valor neutro o simplemente se las excluy\'e.}
%
\en{In Figure~\ref{fig:smooth_example} we show the behavior of the estimates in a data set with two agents and two events.}
\es{En la figura~\ref{fig:smooth_example} se muestra el comportamiento de las estimaciones en conjunto de datos de dos agentes y dos eventos.}
%
\begin{figure}[h!]
  \centering
  \scalebox{.9}{
    \tikz{ %
      \node[latent] (s10) {$s_{a_0}$} ;
      %
      \node[latent,  below=of s10,yshift=-1cm] (s11) {$s_{a_1}$} ;

@@

\en{The efficiency of TrueSkill Throguth Time is due to the fact that margins, whether exact or approximate, are solved analytically.}
\es{La eficiencia de TrueSkill Throguth Time se debe a que las marginales, sean exactas o aproximadas, se resuelven de forma anal\'itica.}
%
\en{In this section we list the properties that we will use to derive the exact messages that arise from the sum-product algorithm.}
\es{En esta secci\'on enumeramos las propiedades que usaremos para derivar los mensajes exactos que surgen del \emph{sum-product algorithm}.}
%
\en{Although these properties are widely known, we attach their full demonstrations in the supplemental material.}
\es{Aunque estas propiedades son ampliamente conocidas, adjuntamos sus demostraciones completas en material suplementario.}

@@

\es{El siguiente mensaje, que la variable $s_i$ le env\'ia al factor rendimiento $f_{p_i}$ tambi\'en será el prior.}

@@

\es{Recordar que los mensajes que env\'ian las variables es tan solo la multiplicaci\'on de los mensajes que esa variable recibe del resto de los factores.}

@@

\es{Por ser de trivial soluci\'on, no haremos expl\'icitos este tipo de mensajes.}
